---
title: "Ecosystem Health Index"
author: "Jessica A. Rowland"
output: html_document
---

## Index overview

The Ecosystem Health Index (EHI) measures temporal changes in environmental conditions and biotic processes/interactions (hereafter collectively, ecological processes). The EHI uses relative severity of change in ecosystem-specific ecological variables and extent of the ecosystem affected to quantify transitions towards or away from ecosystem collapse. The index represents the geometric mean of the relative value of decline. It uses data defined in IUCN Red List of Ecosystems risk assessments.  

This information sheet provides the code used to calculate the the index and an example of each step.  

**Reference:**  
Rowland, J. A., Bland, L. M., Keith, D. A., Bignoli, D. J., Burgman, M., Etter, A., Ferrer-Paris, J. R., Miller, R. M. and Nicholson, E. (2019) Ecosystem indices to support global biodiversity conservation. Conservation Letters. e12680  

## Set up functions

### Calculate the index

The function *calcEHI* selects the column in a dataframe listing the relative severity of degradation and the proportion of the ecosystem area affected by degradation for criteria C (environmental degradation) and D (biotic disruption), combines these into a single value, selected the variable with the highest decline out of criteria C and D, and this value is used to calcalte the index, along with percentiles capturing the middle 95% of the data. 

Parameters are:  

- eco_data = dataframe  
- RLE_criteria = name of the column with the Red List of Ecosystems criterion of interest  
- severity = severity of degradation of an ecological variable over a given timeframe towards ecosystem collapse
- extent = proportion of the ecosystem area affected by degradation
- group1 = the factor (optional) you want to group the index by. Where not specified, an EAI will be calculated based on all ecosystems (output = single score)  
- group2 = the second factor (optional) you want to group the index by  

```{r calcEHI}
calcEHI <- function(eco_data, severity_C, extent_C, severity_D, extent_D, group){
  
  # Calculate proportion remaining for criteria C and D
  funcPct <- mutate(eco_data, 
                    est_intact_C = 1 - (eco_data[[severity_C]] * eco_data[[extent_C]]),
                    est_intact_D = 1 - (eco_data[[severity_D]] * eco_data[[extent_D]])) 
  
  # Select lowest value for C or D
  final <- mutate(funcPct, est_intact = ifelse(is.na(est_intact_C), est_intact_D, 
                                        ifelse(is.na(est_intact_D), est_intact_C,
                                        ifelse(est_intact_C <= est_intact_D, est_intact_C, est_intact_D))))
  # Drop NA values
  no_na_values <- drop_na(final, est_intact)
  
  # Calculate overall index score if missing group, or scores based on a classification if specified
  if (missing(group)) {
    values <- group_by(no_na_values)
    
    } else {
      values <- group_by(no_na_values, group = no_na_values[[group]])
  }
  
  # Calculate index scores (accounting for zeros) and quantiles
  index_scores <- summarise(values, total_count = n(),
                            count_non_zeros = length(est_intact > 0),
                            EHI = ((exp(mean(log(est_intact[est_intact > 0])))) * (count_non_zeros/total_count)),
                            lower = quantile(est_intact, probs = 0.025),
                            upper = quantile(est_intact, probs = 0.975))
}
```

#### Example

This test data set are available from github.

```{r example, message = FALSE}
# Load packages
library(dplyr)
library(tidyr)

# Load data
data <- read.csv("~/Documents/*Career/*PhD/*Projects/2 - RLE indicators/Github script/Github_example_data_EHI.csv")

# View data
head(data)
```

Calculate the index using no groupings:

```{r RLIE overall}
# Calculate the index values
output <- calcEHI(data,
                  severity_C = "C1_severity",
                  extent_C = "C1_extent",
                  severity_D = "D1_severity",
                  extent_D = "D1_extent")

# View output
head(output)
```

Calculate the index using one grouping:

```{r RLIE one group}
# Calculate the index values
output_one_grouping <- calcEHI(data,
                               severity_C = "C1_severity",
                               extent_C = "C1_extent",
                               severity_D = "D1_severity",
                               extent_D = "D1_extent",
                               group = "region")

# View output
head(output_one_grouping)
```

## Author information

jessica.rowland674@gmail.com  
http://jessrowlandresearch.wordpress.com